A simulation model of DNA methylation data

### 2015-04-24 Rmarkdown/methsim.Rmd
Pre-lunch commit. Begin re-write of methsim chapter.

\section{Chapter overview}\label{sec:methsim_overview}

In this chapter I argue for the utility, even necessity, of methodology and software for simulating DNA methylation data. I present the `methsim` software to simulate DNA methylation data. `methsim` incorporates a model of within-fragment co-methylation and simulates individual sequencing reads, which sets it apart from existing software. I describe the statistical model behind `methsim` and its software implementation. Finally, I provide examples of how `methsim` can be used to learn about DNA methylation and how it might be used to develop simulation studies to compare competing analysis methods.

@@

It can be difficult to design an experiment that can be used for validating or benchmarking different analysis strategies. In fact, the huge variety of experimental factors, and their possible values, can make such a task infeasible. But perhaps the bigger hurdle is that there are not attractive experiments to perform; why spend your time and money on an experiment where you "know the answer" when you could be spending that same time and money on investigating something new? In these scenarios and more, simulation studies play a vital role in applied statistics, where they can be used in the development, validation and benchmarking of different analysis methodologies.

@@

Simulation studies can provide many insights into the performance of a method. At the most basic level, if a method fails to work or performs poorly when applied to simulated data, then it is very unlikely to work well when applied to real data. We can also identify which scenarios are "easy", ones where most methods are able to identify the truth, and scenarios under which certain methods perform better than others. While simulated data can never fully capture the richness of "real" data, we can learn a lot about a method by studying how it performs when applied to simulated data.

@@

1. Realism: The simulated data must be "similar" to the real data. While this is obvious, it is also often hard to pin down or agree upon what constitutes "similar enough".
2. Cost: It should be fast and cheap to simulate data. The most common use of simulation models in applied statistics is to repeatedly generate datasets under a range of parameter settings. This requires that each simulation is fast and computationally cheap, otherwise it will be prohibitive to analyse the full space of scenarios. An exception to this rule may occur when the simulation model is used to test a proposed mechanistic or stochastic model of a phenomenon, such as in studies of molecular dynamics. Even then, however, the cost of a simulation should be less than the cost of performing the equivalent experiments, otherwise the simulation is generally not worth the effort.
3. Usability: There __must__ be a software implementation. Simulation models exist to be simulated from; a simulation model without an implementation is next to useless. The implementation should give the user easy access to the key  parameters and have sensible default settings. The output of the software should be in a standard format or readily convertible to a simple, manipulable format.

@@

`methsim` requires an input dataset from which to estimate key parameters. For the input dataset, `methsim` requires the methylation patterns at various sized m-tuples, which can be produced by the `methtuple` software.

This workflow is illustrated in __FIGURE__ and described in the remainder of this section.

@@

`methsim` is written in R and builds on the `MethylationTuples` package described in __CHAPTER__, as well as several R packages available on Bioconductor and CRAN (see `DESCRIPTION` for a complete list. __TODO: Add citations of all packages used in thesis__).

`methsim` makes extensive use of the S4 object system in R. The most important classes defined by `methsim` are `SimulateMethylomeParam`, `SimulatedMethylome`, `SimulateBSParam` and `SimulatedBS`. The most important methods are the `simulate()` methods defined for the `SimulateMethylomeParam` and `SimulateBSParam` classes. The role of these classes and methods in the workflow is illustrated in __FIGURE__.

The `SimulateMethylomeParam` object contains the empirical distributions of key statistics from which the parameters for simulating the 'true' methylome are sampled. Therefore, the `SimulateMethylomeParam` should be based on a relevant sample. To help a new user get started, `methsim` includes `SimulateMethylomeParam` objects for the _ADS_, _ADS-adipose_ and _ADS-iPSC_ samples from the _Lister_ dataset. Alternatively, the user may process the BAM file for their sample with `methtuple` and then use the helper functions in `methsim` to construct a `SimulateMethylomeParam` object based on their sample of interest.

@@

To simulate bisulfite-sequencing reads, the user first constructs a `SimulateBSParam`, which includes the `SimulatedMethylome`. The user then runs the `simulate()` method on the `SimulatedBSParam` object to create a simulated bisulfite-sequencing data in the form of a `SimulatedBS` object.

Finally, we may wish to simplify the `SimulatedBS` object. For example, we may only require the $\beta$-values for our downstream analysis and not the  read-level data. `methsim` includes the `asMethPat()` function to convert the `SimulatedBS` object to a `MethPat` object (described in __CHAPTER__).

__UP TO HERE: See if the remainder of this section needs re-jigging in light of the above changes to Implementation__

@@

As we have seen, DNA methylation is highly heterogeneous along the genome. Nonetheless, there are clear regions of "similarity", such as the unmethylated CpG islands and long partially methylated domains. In these locally-similar regions we might hope to model DNA methylation by a simple parametric model. `methsim` is based on the idea of segmenting the genome into "regions of similarity", fitting a simple model to each region and then "stitching" the results together to form the 'true' methylome.

@@

1. Segment the methylome into regions of similarity.
2. Within each region simulate the binary methylation states from a simple parametric model.
3. "Stitch" the resulting region-level simulations together[^stitch].

[^stitch]: In practice, the stitching together of segments is incorporated into step 2.

Suppose our sample contains $n_{h}$ haplotypes and, therefore, that our sample can be represented by[^strand_symmetric] $\mathbf{Z} = \{ \mathbf{Z_{h}} \}_{h = 1}^{h = n_{h}}$. We might try to simulate this sample by generating $n_{h}$ independent realisations of $\mathbf{Z_{h}}$. However, for a number of reasons, this is a little naive and ambitious. For example, under this naive model there is no control across haplotypes of the methylation patterns; although there exists between-haplotype heterogeneity, the $\mathbf{Z_{h}}$ are definitely not independent of one another.

[^strand_symmetric]: Currently, `methsim` can only simulate CpG methylation and assumes that the methylation states of all CpGs are strand symmetric.

Rather than try to explicitly model dependencies between the $\mathbf{Z_{h}}$, `methsim` uses the following heuristic procedure:

1. Simulate $n_{\tilde{h}} \ll n_{h}$ independent realisations of $\mathbf{{Z_{\tilde{h}}}}$, where $\mathbf{Z_{\tilde{h}}}$ has the same stochastic properties as $\mathbf{Z_{h}}$. `methsim` uses a default value of $n_{\tilde{h}} = 8$.
2. Within each region, $r_{i}$, assign each of $\mathbf{Z_{\tilde{h}}}$ a weight, $W_{r_{i}, \tilde{h}}$, where $\sum_{ \tilde{h} = 1}^{n_{ \tilde{h}}} W_{r_{i}, \tilde{h}} = 1$ and, by construction, $W_{r_{i}, 1} \geq \ldots \geq W_{r_{i}, n_{\tilde{h}}}$.

For simplicity, `methsim` assumes that all $n_{\tilde{h}}$ realisations have the same parameters (discussed below; __TODO: Discuss implications of this assumption somewhere below__).

The weights are interpreted as the proportion of $\{ Z_{h, i} \}_{h = 1}^{h = n_{h}}$ with the methylation pattern given by $Z_{\tilde{h}, i}$ for the $i^{th}$ methylation locus[^weights]. We write the matrix of weights for all methylation loci in the genome as $\mathbf{W} = \{ \mathbf{W_{r_{i}, \cdot}} \}_{i = 1}^{N_{loci}} = \{ \mathbf{W_{\cdot, \tilde{h}}} \}_{\tilde{h} = 1}^{n_{\tilde{h}}}$. In regions of low within-sample heterogeneity, such as those unmethylated regions with $B_{i} \approx 0$, this means we want $\mathbf{W_{r_{i}, \cdot}} \approx (1, 0, \ldots, 0)$, whereas in regions of high within-sample heterogeneity, such as partially methylated regions, we want a more dispersed $\mathbf{W_{r_{i}}, \cdot}$.

[^weights]: Note that these weights are defined at the region-level so all loci in the region have the same weights.

A simulated methylome is therefore described by $\{ \mathbf{{Z_{\tilde{h}}}} \}_{\tilde{h} = 1}^{n_{\tilde{h}}}$ and $\mathbf{W}$. For example, $B_{i} = \sum_{\tilde{h} = 1}^{n_{\tilde{h}}} W_{r_{i}, \tilde{h}} \times Z_{i, \tilde{h}}$.

I now describe each step in this process in greater detail.

@@

`methsim` uses the the R/Bioconductor package, `MethylSeekR` \citep{Burger:2013kq}, to segment the input methylome into these regions of similarity. Initially published to discover regulatory motifs from bisulfite-sequencing data, `MethylSeekR` segments the genome into unmethylated regions (\emph{UMR}s), lowly-methylated regions (\emph{LMR}s) and partially methylated regions[^pmds] (\emph{PMR}s). It does this using a two-stage algorithm applied to the $\beta$-values from a sample:

@@

`methsim` uses the functionality of `MethylSeekR` to partition the genome into "regions of similarity", namely `UMR`, `LMR`, `PMR` or `other`[^other]. While a tailored algorithm may improve the segmentation process, the result produced by `MethylSeekR` is a reasonable approximation to segmenting into "regions of similarity". This can be seen in __FIGURES__  (__TODO: Figures showing that `MethylSeekR` gives a reasonable segmentation for the purposes of `methsim`)__). It should be noted that `methsim` also performs some post-processing of the `MethylSeekR` output so that the resulting segmentation is disjoint and exhaustive.

[^other]: `other` roughly corresponds to 'mostly methylated regions', although \citet{Burger:2013kq} do not describe these regions as such.

__TODO: Figure showing `other` is MMRs. More generally, a figure showing the average methylation level in each of `UMR`, `LMR`, `PMR` and `other`.__

\subsubsection{Simulating $\{ \mathbf{{Z_{\tilde{h}}}} \}_{\tilde{h} = 1}^{n_{\tilde{h}}}$}

`methsim` uses a two-state ($1 =$ methylated or $0 =$ unmethylated), one-step Markov chain to model $\mathbf{{Z_{\tilde{h}}}}$. This allows the incorporation of within-fragment co-methylation into the simulation. The choice of a one-step process is one of computational simplicity, although I will show that this is a reasonable approximation of within-fragment co-methylation.

@@

For these reasons I allow the transition probabilities to vary with $i$ and $r_{i}$, that is, I allow the Markov chain to be spatially inhomogeneous. In particular, I allow the transition probabilities, $p_{i^{(h, j)}, a, b} = Pr(Z_{h, i + 1, j} = b | Z_{h, i, j} = a)$, to depend on the intra-pair distance between the $i^{th}$ and $(i + 1)^{th}$ locus and on the region for this pair of loci[^context], $p_{i^{(h, j)}, a, b} = Pr(Z_{h, i + 1, j} | Z_{h, i, j}, r_{i}, IPD(i, i + 1))$, where $r_{i}$ is the region type (UMR, LMR, PMR or MMR) of the $i^{th}$ methylation locus.

@@

The above-described model, being spatially inhomogeneous, is not particularly amenable to analytical calculations. However, it is very simple to simulate realisations from this model, requiring only a single loop over the set of loci on each chromosome. For a chromosome containing $n$ methylation loci there are $n - 1$ transition matrices, $P_{i^{(h)}}$, to estimate or otherwise assign. Again, I emphasise that `methsim` uses the same set of parameters for all $n_{\tilde{h}}$ "pseudo-haplotypes". Therefore, only one set of transition matrices need to be assigned.

`methsim` assigns these transition probabilities based on $\beta_{i | r_{i}}$ and $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$, the empirical distributions of $\beta_{i}$ and $LOR_{IPD(i, i + 1)}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$, and the intra-pair distance of the the $i^{th}$ and $(i + 1)^{th}$ locus, $IPD(i, i+1)$, respectively. These empirical distributions are shown for the \emph{ADS} sample in Figures \ref{fig:ADS_beta_by_pm_boxplots} and \ref{fig:ADS_lor_by_pm_quantile_plots}.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/ADS_lor_by_pm_quantile_plots.pdf}
\caption{$LOR_{(i, i + 1) | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{ADS} sample. Only 2-tuples with at least $10\times$ sequencing coverage are used. UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
\label{fig:ADS_lor_by_pm_quantile_plots}
\end{figure}

These same distributions for all the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ samples are shown in Figures \ref{fig:EPISCOPE_beta_by_pm_boxplots}, \ref{fig:EPISCOPE_lor_by_pm_quantile_plots}, \ref{fig:Lister_beta_by_pm_boxplots}, \ref{fig:Lister_lor_by_pm_quantile_plots},
\ref{fig:Seisenberger_beta_by_pm_boxplots}, \ref{fig:Seisenberger_lor_by_pm_quantile_plots},
\ref{fig:Ziller_merged_beta_by_pm_boxplots} and \ref{fig:Ziller_merged_lor_by_pm_quantile_plots}.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_by_pm_boxplots.pdf}
\caption{$\beta_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{EPISCOPE} dataset. Only CpGs with at least $10\times$ sequencing coverage are used. UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
\label{fig:EPISCOPE_beta_by_pm_boxplots}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_lor_by_pm_quantile_plots.pdf}
\caption{$LOR_{(i, i + 1) | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{EPISCOPE} sample. Only 2-tuples with at least $10\times$ sequencing coverage are used. UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
\label{fig:EPISCOPE_lor_by_pm_quantile_plots}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_beta_by_pm_boxplots.pdf}
\caption{$\beta_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{Lister} dataset. Only CpGs with at least $10\times$ sequencing coverage are used. UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
\label{fig:Lister_beta_by_pm_boxplots}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_lor_by_pm_quantile_plots.pdf}
\caption{$LOR_{(i, i + 1) | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{Lister} sample. Only 2-tuples with at least $10\times$ sequencing coverage are used. UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
\label{fig:Lister_lor_by_pm_quantile_plots}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_beta_by_pm_boxplots.pdf}
\caption{$\beta_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{Seisenberger} dataset. Only CpGs with at least $10\times$ sequencing coverage are used. UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
\label{fig:Seisenberger_beta_by_pm_boxplots}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Seisenberger_lor_by_pm_quantile_plots.pdf}
\caption{$LOR_{(i, i + 1) | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{Seisenberger} sample. Only 2-tuples with at least $10\times$ sequencing coverage are used. UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
\label{fig:Seisenberger_lor_by_pm_quantile_plots}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_beta_by_pm_boxplots.pdf}
\caption{$\beta_{i | r_{i}}$, the distribution of $\beta$-values for CpGs in each region type, for the \emph{Ziller} dataset. Only CpGs with at least $10\times$ sequencing coverage are used. UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
\label{fig:Ziller_merged_beta_by_pm_boxplots}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_lor_by_pm_quantile_plots.pdf}
\caption{$LOR_{(i, i + 1) | r_{i}}$, the distribution of $\log_{2}$ odds-ratio estimates of within-fragment co-methylation for CpGs in each region type, for the \emph{Ziller} sample. Only 2-tuples with at least $10\times$ sequencing coverage are used. UMR = unmethylated region; LMR = lowly methylated region; PMR = partially methylated region; other = any other region.}
\label{fig:Ziller_merged_lor_by_pm_quantile_plots}
\end{figure}

`methsim` assigns these probabilities as follows:

1. For each region, $r_{i}$, sample $B_{r_{i}}$ from $\beta_{i | r_{i}}$. This can be interpreted as $Pr(Z_{r_{i}} = 1)$, the marginal probability that a methylation locus in region $r_{i}$ is methylated, or, alternatively, as the average level of methylation for all methylation loci in the region[^extreme_B].
2. For every pair of methylation loci, $(i, i + 1)$, $1 \leq i \leq (n - 1)$, sample $\Psi_{(i, i + 1)}$ from $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$. This can be interpreted as the odds that the $(i + 1)^{th}$ methylation locus has the same methylation state as the $i^{th}$ methylation locus.

[^extreme_B]: If $B_{r_{i}} = 0$ (resp. $B_{r_{i}} = 1$), `methsim` adds (resp. subtracts) a small value $\epsilon$. The default value is $\epsilon = 0.01$. The boundary values of $0$ and $1$ are common in $\beta_{i | r_{i}}$ and, were these retained, would result in methylomes with large regions that are uniformly unmethylated (resp. methylated).

The second step requires further explanation. If there are no observations with the required $IPD$-region combination, then, by definition, `methsim` is unable to sample from the appropriate empirical distribution[^min_n]. Instead, when this occurs, `methsim` samples from a user-specified parametric distribution. __RESULTS SECTION__ explores different choices of this distribution __AND SHOWS WHAT__. The default option is a zero-mean Gaussian distribution, chosen so that "on average" the methylation state at adjacent CpGs with large $IPD$s are independent of one another. The default standard deviation of this Gaussian distribution is $2$; this value, while fairly arbitrary, reflects the median standard deviation of the $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$ distributions for each sample, which is remarkably constant across $r_{i}$ and samples[^ipd_stability] (__TODO: Format and reference table__).

[^min_n]: By default, `methsim` requires at least 100 observations in the  $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$ to try to avoid sampling from sparse empirical distributions.

[^ipd_stability]: Figures \ref{fig:EPISCOPE_lor_by_pm_quantile_plots}, \ref{fig:Lister_lor_by_pm_quantile_plots}, \ref{fig:Seisenberger_lor_by_pm_quantile_plots} and \ref{fig:Ziller_merged_lor_by_pm_quantile_plots} show that the variability of these distributions is fairly constant across $IPD$ for a given $r_{i}$ and sample.

```{r, eval = TRUE, echo = FALSE}
knitr::kable(readRDS("../data/median_sd_LOR_distributions.rds"), align = "c", digits = 2)
```

Now that we have $B_{r_{i}}$ and $\Psi_{i, i + 1}$ for $1 \leq i \leq n - 1$ we can simulate realisations of $\mathbf{Z_{\tilde{h}}}$. To begin the process (i.e., when $i = 1$) $Z_{\tilde{h}, 1}$, is simulated from the "marginal" distribution, which is Bernoulli($B_{r_{1}}$). Then, we loop over $i = 2, \ldots, n - 1$ to simulate the remainder of the process. At each step we compute the transition probability, $p_{(i + 1) | i} = Pr(Z_{\tilde{h}, i + 1} = 1 | Z_{\tilde{h}, i}) = \frac{Pr(Z_{tilde{h}, i}, Z_{tilde{h}, i + 1} = 1)}{Pr(Z_{tilde{h}})}$, from $B_{r_{i}}, B_{r_{i + 1}}, \Psi_{(i, i + 1)}$ and the observed (simulated) value at the previous iteration, $Z_{\tilde{h}, i}$. Then, $Z_{\tilde{h}, i + 1}$ is simulated from the Bernoulli($p_{(i + 1) | i}$) distribution.

>>>>

\begin{chapabstract}
I present the `methsim` software to simulate DNA methylation data. `methsim` incorporates a model of within-fragment co-methylation and simulates individual sequencing reads, which sets it apart from existing software. I describe the statistical model behind `methsim` and its software implementation. Finally, I provide examples of how `methsim` can be used to learn about DNA methylation and how it might be used to develop simulation studies to compare competing analysis methods.
\end{chapabstract}

@@

It can be difficult to design an experiment that can be used for validating or benchmarking different analysis strategies. In fact, the huge variety of experimental factors, and their possible values, can make such a task infeasible. But perhaps the bigger hurdle is that there are not attractive experiments to perform; why spend your time and money on an experiment where you 'know the answer' when you could be spending that same time and money on investigating some new biology? In these scenarios and others, simulation studies play a vital role in applied statistics, where they can be used in the development, validation and benchmarking of different analysis methodologies.

@@

Simulation studies can provide many insights into the performance of a method. At the most basic level, if a method fails performs poorly when applied to 'easy' simulated data, then it is very unlikely to work well when applied to more complex real data. We can also identify which scenarios are 'easy' (ones where most methods are able to identify the truth) and scenarios under which certain methods perform better than others. While simulated data can never fully capture the richness of real data, we can learn a lot about a method by studying how it performs when applied to simulated data.

@@

1. Realism: The simulated data must be 'similar' to the real data. While this is obvious, it is also often hard to pin down or agree upon what constitutes 'similar enough'.
2. Cost: It should be fast and cheap to simulate data. The most common use of simulation models in applied statistics is to repeatedly generate datasets under a range of parameter settings. This requires that each simulation is fast and computationally cheap, otherwise it will be prohibitive to analyse the full space of scenarios. An exception to this rule may occur when the simulation model is used to test a proposed mechanistic or stochastic model of a phenomenon, such as in studies of molecular dynamics. Even then, however, the cost of a simulation should be less than the cost of performing the equivalent experiments, otherwise the simulation is generally not worth the effort.
3. Usability: There __must__ be a software implementation. Simulation models exist to be simulated from; a simulation model without an implementation is next to useless. The implementation should give the user easy access to the key  parameters and have sensible default settings. The output of the software should be in a standard format or readily convertible to a simple, manipulable format.

@@

`methsim` requires an input dataset from which to estimate key parameters. For the input dataset, `methsim` requires the methylation patterns at various sized m-tuples, which can be produced by the `methtuple` software. `methsim` can currently only simulate CpG methylation and assumes that the methylation states of all CpGs are strand symmetric.

@@

`methsim` is written in R and builds on the `MethylationTuples` package described in Chapter \ref{chap:wgbs_downstream_analyses}, as well as several R packages available on Bioconductor and CRAN (see the `DESCRIPTION` file for a complete list.).

`methsim` makes extensive use of the S4 object system in R. The most important classes defined by `methsim` are `MethylomeParam`, `SimulatedMethylome`, `WGBSParam` and `SimulatedBS`. The most important methods are the `simulate()` methods defined for the `MethylomeParam` and `WGBSParam` classes.

The `MethylomeParam` object contains the empirical distributions of key statistics from which the parameters for simulating the 'true' methylome are sampled. Therefore, the `MethylomeParam` should be based on data from a relevant sample. To help a new user get started, `methsim` includes `MethylomeParam` objects for the _ADS_, _ADS-adipose_ and _ADS-iPSC_ samples from the _Lister_ dataset. Alternatively, the user may process the `BAM` file for their sample with `methtuple` and then use the helper functions in `methsim` to construct a `MethylomeParam` object based on their sample of interest.

@@

Once we have a 'true' methylome, we can simulate data from it. Currently, `methsim` supports the simulation of whole-genome bisulfite-sequencing reads via the `WGBSParam` class and associated `simulate()` method. Other assays, such as RRBS or microarrays may be supported in future releases. A `WGBSParam` object will contain a `SimulatedMethylome` object, along with parameters such as the read-length, sequencing coverage and error rate of the data to be simulated. When applied to a `WGBSParam` object, which includes a `SimulatedMethylome` object, the `simulate()` method can return a `SimulatedBS` object or a `MethylationTuples::MethPat` object[^namespace2]. A `SimulatedBS` object contains all simulated reads and is generally much larger than a `MethylationTuples::MethPat` object, which summarises the simulated reads for m-tuples of a particular size[^methpat]. `methsim` also includes a helper function, `asMethPat()`, to coerce a `SimulatedBS` object to a `MethylationTuples::MethPat` object.

[^namespace2]: This uses the NAMESPACE notation of R: `MethylationTuples::MethPat` can be read as "the `MethPat` class is part of the `MethylationTuples` package". See Section \ref{sec:MethylationTuples} for details of this class.

[^methpat]: Returning a `MethylationTuples::MethPat` object may be appropriate, for example, if all that is required for downstream analyses are methylation counts at 1-tuples.

@@

As we have seen, DNA methylation is highly heterogeneous along the genome. Nonetheless, there are clear regions of "similarity", such as the unmethylated CpG islands and long partially methylated domains. In these locally-similar regions we might hope to model DNA methylation by a simple parametric model. `methsim` is based on the idea of segmenting the genome into 'regions of similarity', fitting a simple model to each region and then 'stitching' the results together to form the true methylome.

@@

1. Segment the methylome into regions of similarity.
2. Sample parameters for each methylation locus, $i$, based on step 1.

I now describe each step in greater detail.

@@

`methsim` uses the the R/Bioconductor package, `MethylSeekR` \citep{Burger:2013kq}, to segment the input methylome into regions of similarity. `MethylSeekR` was developed to discover regulatory motifs from bisulfite-sequencing data by segmenting the methylome into unmethylated regions (\emph{UMR}s), lowly-methylated regions (\emph{LMR}s) and partially methylated regions[^pmds] (\emph{PMR}s). It does this using a two-stage algorithm applied to the $\beta$-values from a sample:

@@

`methsim` post-processes the segmentation provided by `MethylSeekR` to partition the methylome into regions of similarly, namely \emph{UMR}s, \emph{LMR}s, \emph{PMRS}s and \emph{other}s[^methylseekr_output]. Roughly speaking, _other_ regions are 'mostly methylated regions' (see Figures \ref{fig:EPISCOPE_beta_by_pm_boxplots}, \ref{fig:Lister_beta_by_pm_boxplots}, \ref{fig:Seisenberger_beta_by_pm_boxplots}, \ref{fig:Ziller_beta_by_pm_boxplots}), although \citet{Burger:2013kq} do not describe these regions as such.

[^methylseekr_output]: The output returned by `MethylSeekR` does not strictly partition the methylome since it is neither disjoint nor exhaustive.

While a tailored algorithm may improve this segmentation process, the result produced by `MethylSeekR` is a reasonable approximation to segmenting the methylome into the required 'regions of similarity'.

\subsubsection{Parameterising the model}

`methsim` uses a two-state ($1 =$ methylated or $0 =$ unmethylated), one-step Markov chain to model $\mathbf{Z} = \{ \mathbf{Z_{h}} \}_{h = 1}^{h = n_{h}}$. This allows the incorporation of within-fragment co-methylation into the simulation. The choice of a one-step process is one of computational simplicity, although I will show that this is a reasonable approximation of within-fragment co-methylation.

@@

For these reasons I allow the transition probabilities, $\mathbf{p}$, to vary with $i$, that is, I allow the Markov chain to be spatially inhomogeneous. In particular, I allow the transition probabilities to depend on the intra-pair distance between the $i^{th}$ and $(i + 1)^{th}$ locus and on the region type, $r_{i}$ (_UMR_, _LMR_, _PMR_ or _other_), for this pair of loci[^context]. That is, $p_{i; a, b} = Pr(Z_{h, i + 1} = b | Z_{h, i} = a, IPD(i, i + 1), r_{i})$.

@@

The above-described model is not particularly amenable to analytical calculations due to the spatial inhomogeneity. It is, however, relatively simple to simulate realisations from this model, requiring only a single loop over the set of loci on each chromosome. For a chromosome containing $n$ methylation loci, there are $n - 1$ transition matrices to estimate or otherwise assign.

Rather than directly modelling the transition probabilities, $\mathbf{p}$, `methsim` is parameterised $\mathbf{Z}$ by a vector of marginal probabilities, $\mathbf{B} = \{ Pr(Z_{i} = 1) \}$, and a vector of odds ratios, $\mathbf{\psi} = \{ \frac{Pr_{Z_{i + 1} = 1 | Z_{i} = 1} \times Pr_{Z_{i + 1} = 0 | Z_{i} = 0}}{Pr_{Z_{i + 1} = 1 | Z_{i} = 0} \times Pr_{Z_{i + 1} = 0 | Z_{i} = 1}} \} = \{ \frac{Pr_{Z_{i} = 1, Z_{i + 1} = 1} \times Pr_{Z_{i} = 0 , Z_{i + 1} = 0}}{Pr_{Z_{i} = 1 , Z_{i + 1} = 0} \times Pr_{Z_{i} = 0 , Z_{i + 1} = 1}} \}$. We can compute the transition probabilities, $\mathbf{p}$, from $\mathbf{B}$ and $\mathbf{\psi}$ as I now describe.

For each 2-tuple, $(i, i + 1)$, `methsim` first constructs the joint probability matrix, $P_{i, i + 1}$ from the marginal probabilities, $B_{i}, B_{i + 1}$, and the odds ratio, $\psi_{i}. The general form of $P_{i, i + 1}$ is shown below:

\begin{equation*}
P_{i, i + 1} = \bordermatrix{~ & 1 - B_{i+1} & B_{i+1} \cr
1 - B_{i} & Pr(Z_{i} = 0, Z_{i + 1} = 0) & Pr(Z_{i} = 0, Z_{i + 1} = 1) \cr
B_{i} & Pr(Z_{i} = 1, Z_{i + 1} = 0) & Pr(Z_{i} = 1, Z_{i + 1} = 1) \cr}
\end{equation*}

`methsim` computes $P_{i, i + 1}$ using the iterative proportional fitting algorithm. Iterative proportional fitting is a general method "for constructing tables of numbers satisfying certain constraints" \citep{Speed:2005tg}. In the case of `methsim`, we use iterative proportional fitting to construct the _unique_ $2 \times 2$ array (the joint probability matrix) with specified margins (the marginal probabilities of each methylation state at each locus) and the specified cross-ratio (the odds-ratio). An example of this procedure is illustrative.

Let $B_{i} = 0.7$, $B_{i + 1} = 0.6$ and $\psi_{i, i + 1} = 2$. To begin the iterative proportional fitting algorithm, form the matrix $P_{i, i + 1}^{(0)} = \Bigl(\begin{smallmatrix}
\psi & 1 \\
   1 & 1
\end{smallmatrix} \Bigr)$. This matrix has the desired cross-ratio, $\psi$, but not the desired row and column margins. At each iteration of the algorithm, iterative proportional fitting adjusts the rows and columns such that the cross-ratio remains $\psi$ while the row and column margins converge towards the desired values. The algorithm continues in this manner, forming a series of $2 \times 2$ tables that converge pointwise (and uniquely) to a $2 \times 2$ table $P_{i, i + 1}^{(*)} \equiv P_{i, i + 1}$. For our example, $P_{i, i + 1}^{(*)} = \Bigl(\begin{smallmatrix}
0.155 & 0.145 \\
0.245 & 0.455
\end{smallmatrix} \Bigr)$, to three decimal places of precision. It is easily verified that $P^{(*)}$ has the desired cross-ratio and marginal sums.

Once $P_{i, i + 1}$ is computed, the desired transition probability is obtained by dividing the appropriate element of $P_{i, i + 1}$ by the appropriate marginal probability. To continue our example, suppose $Z_{i} = 0$, then the probability that $Z_{i + 1}$ is also to zero is given by $Pr(Z_{i + 1} = 0 | Z_{i} = 0) = \frac{Pr(Z_{i = 0}, Z_{i + 1} = 0)}{Pr(Z_{i} = 0)} = \frac{0.155}{0.3} = 0.517$ to three decimal places. In fact, `methsim` only stores $Pr(Z_{i + 1} = 1 | Z_{i} = 0)$ and $Pr(Z_{i + 1} = 1 | Z_{i} = 1)$ since Pr(Z_{i + 1} = 0 | Z_{i} = 0) = 1 - Pr(Z_{i + 1} = 1 | Z_{i} = 0)$ and $Pr(Z_{i + 1} = 0 | Z_{i} = 1) = 1 - Pr(Z_{i + 1} = 1 | Z_{i} = 1)$.

To simulate values of $\mathbf{B}$ and $\mathbf{\psi}$, `methsim` uses the empirical distributions of their estimators, $\mathbf{\beta} and $\mathbf{\widehat{\psi}}$, in the input methylome. More specifically, $\mathbf{B}$ is based on $\beta_{i | r_{i}}$ and $\mathbf{\psi}$ is based on $\widehat{\psi_{IPD(i, i+1) | r_{i}}$, the empirical distributions of $\beta_{i}$ and $\psi_{IPD(i, i + 1)}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$, and the intra-pair distance of the the $i^{th}$ and $(i + 1)^{th}$ locus, $IPD(i, i+1)$, respectively. These empirical distributions are shown for the \emph{ADS} sample in Figures \ref{fig:ADS_beta_by_pm_boxplots} and __FIGURE__.

@@

__TODO: Figure of ADS within-fragment co-methylation__

These same distributions for all the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ samples are shown in __FIGURES__.

There are several ways to simulate values of $\mathbf{B}$ and $\mathbf{\psi}$ based on these empirical distributions. The simplest is to sample each $B_{i}$ and $\psi_{i}$ directly from these empirical distributions. Alternatively, we could take a more model-based approach. For example, we could specify a  parametric model of the $B_{i}$ where all $i$ in a region have the same mean and variance (the mean and variance being based on the empirical distribution of $\beta$-values) or where the $B_{i}$ in each region come from an autoregressive process.

The simulation of $\mathbf{\psi}$ requires further explanation. The empirical distributions of $\mathbf{\widehat{\psi}}$ will only include values for small $IPD$s since these are estimated from individual sequencing reads (see Chapter \ref{chap:co-methylation} for details). When simulating a methylome, we will encounter 2-tuples with much larger $IPD$s and the question is how to deal with these? Again, there are several options. We might assume a parametric form for the distribution of $\mathbf{\psi}$ as a function of $IPD$, such as a exponential decay towards $\psi = 1$ (corresponding to independence).

I explore different methods for simulating $\mathbf{B}$ and $\mathbf{\psi}$ in Section \ref{sec:results}.

The above-described model implicitly assumes that all $n_{h}$ haplotypes have the same marginal probabilities and co-methylation structure. We may extend this model to define $\mathbf{Z}$ as a _mixture_ of a small number of first-order Markov chains defined as in the above. Such a mixture model may be appropriate when simulating a sample that is a combination of cell types.
